Characteristic polynomials of p-adic matrices

### 2017-01-30 charpoly/charpoly.tex
Computation in O~(n^omega) of Com(X-M)

@@

Let $C$ be the companion matrix associated to $\chi_M$:
$$N = \left( \begin{matrix}
0 & 1 & 0 & \cdots & 0 \\
\vdots & \ddots & 1 & \ddots & \vdots \\
\vdots & & \ddots & \ddots & 0 \\
0 & \cdots & \cdots & 0 & 1 \\
-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
\end{matrix} \right)$$
with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
By Proposition~\ref{prop:surjectivity}, there exists a matrix 
$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
result to the transpose of $M$, we find that there exists another
invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.

\begin{prop}
\label{prop:shortcom}
We keep the previous notations and assumptions.
Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
$$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
\mod \chi_M$$
for some $\alpha \in K[X]$.
\end{prop}

@@

>>>>

\newcommand{\inv}{\text{\rm inv}}

@@

Let $C$ be the companion matrix associated to $\chi_M$:
\begin{equation}
\label{eq:companion}
N = \left( \begin{matrix}
0 & 1 & 0 & \cdots & 0 \\
\vdots & \ddots & 1 & \ddots & \vdots \\
\vdots & & \ddots & \ddots & 0 \\
0 & \cdots & \cdots & 0 & 1 \\
-a_0 & -a_1 & \cdots & \cdots & -a_{n-1}
\end{matrix} \right)
\end{equation}
with $\chi_M = a_0 + a_1 X + \cdots + a_{n-1} X^{n-1} + X^n$.
By Proposition~\ref{prop:surjectivity}, there exists a matrix 
$P \in \GL_n(K)$ such that $M = P N P^{-1}$. Applying the same
result to the transpose of $M$, we find that there exists another
invertible matrix $Q \in \GL_n(K)$ such that $M^t = Q N Q^{-1}$.

\begin{prop}
\label{prop:shortcom}
We keep the previous notations and assumptions.
Let $V$ be the row vector $(1, X, \ldots, X^{n-1})$. Then
\begin{equation}
\label{eq:shortcom}
\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t
\mod \chi_M
\end{equation}
for some $\alpha \in K[X]$.
\end{prop}

@@

We now move to algorithms and show that the quadruple $(\alpha, P, Q, 
\chi_M)$ can be computed for a cost of $\softO(n^\omega)$ operations
in the ground field $K$ with high probability.

The matrix $P$ can be computed as follows. Pick $c \in K^n$. Define 
$c_i = M^i c$ for all $i \geq 1$. The $c_i$'s can be computed in 
$\softO(n^\omega)$ operations in $K$ using \cite{}. Let $P_\inv$ be the 
$n \times n$ matrix whose rows are the $c_i$'s for $1 \leq i \leq n$. 
Remark that $P_\inv$ is invertible if and only if $(c_0, c_1, \ldots, 
c_{n-1})$ is a basis of $K^n$ if and only if $c$ is a cyclic vector. 
Moreover after base change to the basis $(c_0, \ldots, c_{n-1})$, the matrix 
$M$ takes the shape \eqref{eq:companion}. In other words, if $P_\inv$
is invertible, then $P = P_\inv^{-1}$ is a solution of $M = P N P^{-1}$.
Observe moreover that the condition ``$P_\inv$ is invertible'' is open
for the Zariski topology. It then happens with high probability as soon
as it is not empty, that is as soon as $M$ admits a cyclic vector, which
holds by assumption.

The characteristic polynomial $\chi_M$ can be recovered thanks to the
relation $(a_0, a_1, \ldots, a_{n-1}) = -c_n \cdot P$.

Now, instead of computing directly $Q$, we first compute a matrix $R$ 
with the property that $N^t = R N R^{-1}$. For that, we apply the same
strategy as above except that we start with the vector $e = (1, 0, 
\ldots, 0)$ (and not with a random vector). A simple computation shows
that, for $1 \leq i \leq n{-}1$, the vector $N^i e$ has the shape:
$$N^i e = (0, \ldots, 0, -a_0, \star, \ldots, \star)$$
with $n{-}i$ starting zeros. Therefore the $N^i e$'s form a basis of
$K^n$, \emph{i.e.} $e$ is always a cyclic vector of $N$. Once $R$ has
been compute, we recover $Q$ using the relation $Q = P_\inv^t R$.

It remains to compute the scaling factor $\alpha$. For this, we write
the relation:
\begin{equation}
\label{eq:shortcomN}
\com(X{-}N) = \alpha \cdot V^t \cdot V R^t \mod \chi_M
\end{equation}
which comes from Eq.~\eqref{eq:shortcom} after multiplication on the 
left by $P^{-1}$ and multiplication on the right by $P$. We observe
moreover that the first row of $R$ is $(1, 0, \ldots, 0)$. Evaluating
the top left entry of Eq.~\eqref{eq:shortcomN}, we end up with the 
relation:
$$\alpha = a_1 + a_2 X + \cdots + a_{n-1} X^{n-2} + X^{n-1}.$$
No further computation are then needed to derive the value of $\alpha$.
